HiC Contact Map Comaprison Using Graphlet Approach

### 2018-04-18 papers/manuscripts/manuscript.tex
Modified DL graph in tensorflow; improved manuscript

Result of pair-wise contact map graphlet distances is
illustrated in Figure \ref{fig:orbit-distances_all}.
Each point on the graph is the
average of the graphlet distance vector 
of the two cell lines specified in the legend
($\bar{\mathbf{d}}^{\scriptscriptstyle A,B}_{i,j}$).
 A one sided paired t-test
was conducted in order to compare distances to
zero and also to each other. 
The resulting p-values showed highly 
significant differeces from zero for all pairs of
cells. The results of tests for $C_{1,1}$ 
is shown in table \ref{tab:t-test}
Results of all t-tests for each
contact map can be found in supplementary materials.
For each contact map, each pair of cells are
ordered based on whether they are statistially
larger that the other or not with a signifiance
of $0.01$.
\begin{table}[]
    \centering
    \resizebox{.5\textwidth}{!}{
    \begin{tabular}{l|ccccccc}
        \rule{0pt}{25pt}
                  & One       & MIT-RL       & ALL-RL       & ALL-MIT      & ALL-CALL4    & CALL4-MIT    & CALL4-RL     \\ \hline
        \rule{0pt}{25pt}
        One       & -         & \textgreater & \textgreater & \textgreater & \textgreater & \textgreater & \textgreater \\
        \rule{0pt}{25pt}
        MIT-RL    & \textless & -            & \textless    & \textless    & \textless    & \textless    & \textless    \\
        \rule{0pt}{25pt}
        ALL-RL    & \textless & \textgreater & -            & \textless    & \textless    & \textless    & \textless    \\
        \rule{0pt}{25pt}
        ALL-MIT   & \textless & \textgreater & \textgreater & -            & =            & =            & =            \\
        \rule{0pt}{25pt}
        ALL-CALL4 & \textless & \textgreater & \textgreater & =            & -            & =            & =            \\
        \rule{0pt}{25pt}
        CALL4-MIT & \textless & \textgreater & \textgreater & =            & =            & -            &              \\
        CALL4-RL  & \textless & \textgreater & =            & =            & =            & =            & -           
        \rule{0pt}{25pt}
    \end{tabular}}
    \caption{
        Result of one-sided t-test for intrachromosomal
        contact map of chromosome 1.
        $\mathbf{d}^{\scriptscriptstyle CALL4, MIT}_{1,1}$ 
        is statistically larger than 0,
        but less than $\mathbf{d}^{\scriptscriptstyle ALL, RL}_{1,1}$ 
        and so forth. We can also conclude that 
        $\mathbf{d}^{\scriptscriptstyle ALL, CALL4}_{1,1}$ is not
        statistically different from 
        $\mathbf{d}^{\scriptscriptstyle ALL, MIT}_{1,1}$
        between graphlets extracted from all contact maps. 
        Refer to supplementary
        material for results of all hypotheis tests.
    }
    \label{tab:t-test}
\end{table}

We caclulated pair-wise MIC for each orbit in each of the 276
contact maps from MIT data and ALL, RL and CALL4 data separately. 
Figure \ref{fig:contact_maps_correlations_all}
shows average orbit correlations across all contact maps, while
figure \ref{fig:orbits_correlations_all} 
demonstrates average correlations across all 72 orbits within each  contact map.
It is worth mentioning that interchromosomal thresholded contact maps 
represent
a bipartide graph with the loci from each chromosome on one side. Due to this
bipartide nature of the graphs in inter-chromosomal maps,
count of certain orbits is always 0, resulting in
a correlation values of 0 for them as well.
We ignored these values  when we calculated averages across orbits 
in figure \ref{fig:contact_maps_correlations_all} since they
would result in a bias towards zero in averages. You can see the bias in 
figure \ref{fig:orbits_correlations_all} where average correlations of orbits
$\mathbb{Q} = \{3, 9, 10{\text -}14, 20{\text -}34, 39{\text -}48, 51{\text -}72\}$ 
are close to zero. In fact all correlations
corresponding to these orbits are 0 except for the ones between the same 
chromosomes.

Figure \ref{fig:orbits_correlations_all} clearly illustrates certain
orbits of Leukemic have higher correlation to each other than to the
normal MIT cell. In fact our statistical analysis shows that 
\textit {for orbits in $\mathbb{Q}$, 
intra-leukemic orbit correlations are significantly higher
than leukemic-normal orbit correlations}. This implies
there are significant differences between normal and
leukemic cells in terms of their local structure.
\begin{figure}
    \centering
    \begin{subfigure}[b]{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_all.png}
        \caption{}
        \label{fig:orbit-distances_all}
    \end{subfigure}
    \begin{subfigure}[b]{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/contact_maps_correlations_all.png}
        \caption{}
        \label{fig:contact_maps_correlations_all}
    \end{subfigure}
    \begin{subfigure}[b]{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/orbits_correlations_all.png}
        \caption{}
        \label{fig:orbits_correlations_all}
    \end{subfigure}
    \caption{   
        \textbf{(a) Pair-wise graphlet signature difference for
        all 276 contact maps:}
        Each point on a graph is the result of averaging all
        the distances across all loci of that contact map.
        ($\bar{\mathbf{d}}^{\scriptscriptstyle A,B}_{i,j} \quad
        \forall i,j \in \{1 ... 23\} \quad \& \quad j \ge i$).  
        \vspace{.1cm} \\
        \textbf{(b) pair-wise average contact map orbit correlations
         for all contact maps:}
        ($\bar{\mathbf{m}}^{\scriptscriptstyle \scriptscriptstyle A,B}_{i,j} 
        \quad \forall i,j \in \{1 ... 23\} \quad \& \quad j \ge i$:
        average along the red \textit{vertical} arrow in figure 
        \ref{graphlet_correlation_schema}).
        These values are calculated by averaging over 
        pairwise correlations of orbits of
        $\mathbb{Q}$ in a contact map.
        \vspace{.1cm} \\
        \textbf{(c) pair-wise average orbit correlations:}
        In figure \ref{fig:orbits_correlations_all}, each point
        in the graph is the result of averaging pair-wise
        orbit correlations over all contact maps
        ($\frac{1}{276}\sum_{i=0}^{23}\sum_{j=i}^{23}{
        m^{\scriptscriptstyle \scriptscriptstyle A,B}_{i,j, o}} \quad 
        \forall o \in \{0, 1, ..., 72\}$:
        average along the red \textit{horizontal} arrows in figure 
        \ref{graphlet_correlation_schema}).
        Counts for certain orbits are always zero in inter-chromosomal
        maps, leading to average value close to zero in 
        Figure \ref{fig:orbits_correlations_all}.
     }
    \label{fig:results_all}
\end{figure}

We have conducted one-sided
t-test in order to test whether the average correlations across
contact maps is equal to 1  and whether the average 
correlations across orbits is equal to 1. The results
for both test showed that all values are significantly less than 1.
Please refer to supplementary material for result of the 
full list of t-test results.
\newpage
\section{Resources}
\textbf{Hi-C Datasets:}
\begin{enumerate}
    \item \href{https://github.com/rasoolianbehnam/watson}{Code base for this article}
    \item \href{http://sysbio.rnet.missouri.edu/T0510/tmp_download/link_to_download_genome_data/}
        {Datasets including cancerous cells}
    \item \href{https://bcm.app.box.com/v/aidenlab/folder/11234760671}{Original Datasets}
\end{enumerate}

>>>>

\begin{figure*}
    \centering
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_chr14.png}
        \caption{}
        \label{fig:orbit-distances_chr14}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/contact_maps_correlations_chr14.png}
        \caption{}
        \label{fig:contact_maps_correlations_chr14}
    \end{subfigure}
    \caption{   
        \textbf{(a) Pair-wise graphlet signature difference for
        all 276 contact maps:}
        Each point on a graph is the result of averaging all
        the distances across all loci of that contact map.
        ($\bar{\mathbf{d}}^{\scriptscriptstyle A,B}_{i,j} \quad
        \forall i,j \in \{1 ... 23\} \quad \& \quad j \ge i$).  
        \vspace{.1cm} \\
        \textbf{(b) pair-wise average contact map orbit correlations
         for all contact maps:}
        ($\bar{\mathbf{m}}^{\scriptscriptstyle \scriptscriptstyle A,B}_{i,j} 
        \quad \forall i,j \in \{1 ... 23\} \quad \& \quad j \ge i$:
        average along the red \textit{vertical} arrow in figure 
        \ref{graphlet_correlation_schema}).
        These values are calculated by averaging over 
        pairwise correlations of orbits of
        $\mathbb{Q}$ in a contact map.
     }
    \label{fig:results_all}
\end{figure*}

Result of pair-wise contact map graphlet distances is
illustrated in supplementary materials. 
As an example, we plotted average
distances only for chromosome 
14 in Figure \ref{fig:orbit-distances_chr14}.
Each point on the graph is the
average of the graphlet distance vector 
of the two cell lines specified in the legend
($\bar{\mathbf{d}}^{\scriptscriptstyle A,B}_{i,j}$).

By comparing signature distance vectors, one can find
how contact maps
differ from each other in terms of local structure. 
Contact maps can serve as measures of spatial proximity between
loci. Graphlets capture certain patterns of interaction, or
in other words, spatial neighborhood for each loci. Thus, if
signature vectors of two loci are close,
 it can be inferred that they have 
similar spatial neighborhood.

As is shown in
figure \ref{fig:orbit-distances_chr14} as well as t-test results
(to be discussed next in this paper),
we can compare pairs of contact maps in terms of their closeness
to each other. 
As an example, our results show that
for most inter-chromosomal contact maps, ALL
and RL cell lines are closer to each other than to other
cell lines. as  shown in 
\ref{fig:orbit_distances_chr14}, this observation is
true for the first 13 inter-chromosomal 
contact maps of chromosome 14 as well as it intra-chromosomal
contact map.
Although comparisons can be made for each contact map as to
whether which contact maps are more similar to each other,
no global pattern has been observed from graphlet distances.

As an alternative, we can compare graphlets by calculating
how often certain graphlets occur in them. By doing so
we measure the frequency distribution
of certain spatial structures in
each contact map. As a result of this approach
we can compare contact maps by calculating the 
correlation between their orbit distributions.
A higher correlation would mean higher similarity
in terms of spatial structure between the loci
involved.

We caclulated pair-wise MIC values 
for each orbit in each of the 276
contact maps from MIT, ALL, RL, and CALL4 data separately. 
As an example,
figure \ref{fig:contact_maps_correlations_chr14}
shows average orbit correlations across contact maps
for chromosome 14.
It corroborates the
concolusion we previously made that ALL and RL cell lines 
are more similar to each other than to other pairs by showing
significanly larger (refer to supplementary material for statistical tests)
orbit correlations between contact maps.
It is worth mentioning that interchromosomal thresholded contact maps 
represent
a bipartide graph with the loci from each chromosome on one side. Due to this
bipartide nature of the graphs in inter-chromosomal maps,
count of certain orbits is always 0, resulting in
a correlation values of 0 for them as well.
We ignored these values  when we calculated averages across orbits 
in figure \ref{fig:contact_maps_correlations_chr14} since they
would result in a bias towards zero in averages. You can see the bias in 
figure \ref{fig:orbits_correlations_all} where average correlations of orbits
$\mathbb{Q} = \{3, 9, 10{\text -}14, 20{\text -}34, 39{\text -}48, 51{\text -}72\}$ 
are close to zero. In fact all correlations
corresponding to these orbits are 0 except for the ones between the same 
chromosomes.

This approach also allows for comparison of orbits
themsleves. By comparing average oribit correlations
across all contact maps of two cell lines, we can
have a measure of how similar certain orbits are
between cell lines. This can give us a 
\textit{global} measure
of how similar two cell lines are in terms of 
\textit{local} spatial positionings.
Figure \ref{fig:orbits_correlations_all} 
demonstrates average correlations across all 72 orbits within each contact map.
Figure \ref{fig:orbits_correlations_all} clearly illustrates certain
orbits of Leukemic have higher correlation to each other than to the
normal MIT cell. In fact our statistical analysis shows that 
\textit {for orbits in $\mathbb{Q}$, 
intra-leukemic orbit correlations are significantly higher
than leukemic-normal orbit correlations}. This implies
there are significant differences between normal and
leukemic cells in terms of their local structure.

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/orbits_correlations_all.png}
    \caption{
        \textbf{Pair-wise average orbit correlations:}
        In figure \ref{fig:orbits_correlations_all}, each point
        in the graph is the result of averaging pair-wise
        orbit correlations over all contact maps
        ($\frac{1}{276}\sum_{i=0}^{23}\sum_{j=i}^{23}{
        m^{\scriptscriptstyle \scriptscriptstyle A,B}_{i,j, o}} \quad 
        \forall o \in \{0, 1, ..., 72\}$:
        average along the red \textit{horizontal} arrows in figure 
        \ref{graphlet_correlation_schema}).
        Counts for certain orbits are always zero in inter-chromosomal
        maps, leading to average value close to zero in 
        Figure \ref{fig:orbits_correlations_all}.
        In this figure, cancer cells data points are depicted
        in warm colors while normal cells are depicted in
        cold colors in increased contrast. As can be seen
        orbit distributions are more similar to each other
        for cancer cells.
    }
    \label{fig:orbits_correlations_all}
\end{figure*}

\subsection{Statistical Analysis}
A one sided paired t-test
was conducted in order to compare pair-wise distances to
zero and also to each other. 
The resulting p-values showed highly 
significant differeces from zero for all pairs of
cells. The results of tests for $C_{1,1}$ 
is shown in table \ref{tab:t-test}
Results of all t-tests for each
contact map can be found in supplementary materials.
For each contact map, each pair of cells are
compared based on whether they are statistially
larger that the other or not with a signifiance
of $0.01$.

\begin{table}[]
    \centering
    \resizebox{.5\textwidth}{!}{
    \begin{tabular}{l|ccccccc}
        \rule{0pt}{25pt}
                  & One       & MIT-RL       & ALL-RL       & ALL-MIT      & ALL-CALL4    & CALL4-MIT    & CALL4-RL     \\ \hline
        \rule{0pt}{25pt}
        One       & -         & \textgreater & \textgreater & \textgreater & \textgreater & \textgreater & \textgreater \\
        \rule{0pt}{25pt}
        MIT-RL    & \textless & -            & \textless    & \textless    & \textless    & \textless    & \textless    \\
        \rule{0pt}{25pt}
        ALL-RL    & \textless & \textgreater & -            & \textless    & \textless    & \textless    & \textless    \\
        \rule{0pt}{25pt}
        ALL-MIT   & \textless & \textgreater & \textgreater & -            & =            & =            & =            \\
        \rule{0pt}{25pt}
        ALL-CALL4 & \textless & \textgreater & \textgreater & =            & -            & =            & =            \\
        \rule{0pt}{25pt}
        CALL4-MIT & \textless & \textgreater & \textgreater & =            & =            & -            &              \\
        CALL4-RL  & \textless & \textgreater & =            & =            & =            & =            & -           
        \rule{0pt}{25pt}
    \end{tabular}}
    \caption{
        Result of one-sided t-test 
        with significance $\alpha = 0.01$
        for intrachromosomal
        contact map of chromosome 1.
        As an example interpretation,
        $\mathbf{d}^{\scriptscriptstyle CALL4, MIT}_{1,1}$ 
        is statistically larger than 0,
        but less than $\mathbf{d}^{\scriptscriptstyle ALL, RL}_{1,1}$ . 
        An equal sign actually means that the two vectors are
        not statistically different from each other.
        We can also conclude that 
        $\mathbf{d}^{\scriptscriptstyle ALL, CALL4}_{1,1}$ is not
        statistically different from 
        $\mathbf{d}^{\scriptscriptstyle ALL, MIT}_{1,1}$
        between graphlets extracted from all contact maps. 
        Refer to supplementary
        material for results of all hypotheis tests.
    }
    \label{tab:t-test}
\end{table}
We have conducted one-sided
t-test in order to test whether the average correlations across
contact maps is equal to 1  and whether the average 
correlations across orbits is equal to 1. The results
for both test showed that all values are significantly less than 1.
Please refer to supplementary material for result of the 
full list of t-test results.

\section{Resources}
\textbf{Hi-C Datasets:}
\begin{enumerate}
    \item \href{https://github.com/rasoolianbehnam/watson}{Code base for this article}
    \item \href{http://sysbio.rnet.missouri.edu/T0510/tmp_download/link_to_download_genome_data/}
        {Datasets including cancerous cells}
    \item \href{https://bcm.app.box.com/v/aidenlab/folder/11234760671}{Original Datasets}
\end{enumerate}
